\documentclass[12pt]{article}
\renewcommand*\familydefault{\sfdefault}
%\usepackage{amssymb,amsmath,amsfonts,comment}
%\usepackage{amsmath,amssymb,graphicx,subfigure,psfrag}
\usepackage{amsmath,amssymb,graphicx,subfigure,psfrag}
\usepackage{amssymb,mathrsfs}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{color,pdfcolmk}
\newcommand{\todo}[1]{\noindent\emph{\textcolor{red}{Todo: #1\:}}}
\newcommand{\note}[1]{\noindent\emph{\textcolor{green}{Note: #1\:}}}
\newcommand{\referee}[1]{\vspace{3ex}\noindent{\textcolor{blue}{#1}}\\[2ex]}
\newcommand{\nnn}{\mathbf{n}}
\newcommand{\fff}{\mathbf{f}}
\newcommand{\uuu}{\mathbf{u}}
\title{Reply to reviewer 1}

\begin{document}

\subsection*{Reply to Referee 1:}

Thanks for the careful reading and the helpful comments.
Please find below point-by-point replies to your comments and
questions. To give you an overview of all the changes in the paper, we
also provide a diff-document that highlights the changes between the
initial submission and this resubmission.

\subsubsection*{Detailed replies (your comments in blue)}

\referee{
Summary impression:\\[.2ex]
Given the interest in high-order methods and the notable cost of the
underlying solver for poisson-type problems in addition to the variety
of options, the core thesis of the manuscript is well-placed.  Yet,
the presentation of the numerics does not lead the reader to conclude
anything from the comparison --- ie, there are too many ways to
interpret the convergence based on iterations and the cost of the
methods.  The primary concern as stated in the Contributions section
was to address the "actual performance", but as it stands the results
do not attempt to answer this.  Indeed, actual timings are avoided
(and clearly stated this way), the performance model is not realistic
(also as stated, since it can vary from implementation and
architecture), and the iterations are not combined with the perceived
cost in a clear way.  Again, the targeted contributions of this
manuscript would be quite interesting, but the numerics and discussion
could be significantly improved to support many of the statements.}
\noindent
Our ``performance metric'' is not the solution time, which strongly
depends on the architecture and the implementation, but the number of
fine-grid MatVecs required for convergence. These MatVecs are the main
building block of all algorithm variants compared in our manuscript as
they dominate the computation both in terms of memory reads/writes and
floating point operations. Thus, our performance model only counts the
number of fine-grid MatVecs. The actual time-to-solution is---at first
order---the number of required MatVecs multiplied by the
time-per-MatVec, which strongly depends on architecture and
implementation. In the revised manuscript, we have tried to better
work out and explain that difference, and we added references to
papers that study the time-per-MatVec and compare matrix-fee versus
assembled MatVecs on specific architectures for hexahedral elements
\cite{ptatin3d,BursteddeGhattasGurnisEtAl08,DevilleFischerMund02}.
% \todo{Incorporate that in paper; should we use a different term instead of ``performance''?}

\referee{
Non-Minor Concerns\\[2ex]
It's likely that h, p, or structured low-order MG are preferred to a
pure AMG method, but the restricition to "discontinuous or anisotropic
coefficients, nor consider ill-shaped elements" on page 3 should
really consider the implications of this.  The three methods will be
challenged in this setting whereas AMG has been used effectively.}
\noindent
We are not trying to make the point that $h$ or $p$ geometric
multigrid is superior to AMG. However, note that the limitations we
emphasize are not fundamental limitations of geometric MG, but we are
trying to be explicit about which test problems we consider in this
paper, and which \emph{not}. 
% \note{We have added language to the manuscript to clarify the advantages and robustness of AMG.}


\referee{ There is a consistent claim that matrix assembly for
  high-order should be avoided, but some context or citation is needed
  to support this (very) general remark.  If the coefficients and
  elements vary widely then the cost of reconstruction may benefit
  from a matrix assembly, particularly if it is repeatedly used.  From
  the results it's not clear at all that a SpMV based on an assembled
  matrix is *not* the most efficient action.  In the end was AMG
  tested?  THe last line appears that it was: "When combined with
  algebraic multigrid for the low-order operator, the smoother on the
  finest mesh can either use the low-order or the high-order
  residual. Initial numerical tests indicate that the latter choice is
  advantageous, but this should be studied more systematically."
  Perhaps this is referring to some side tests; it is suggested then
  to support these statements more.  Also, the code appears to use the
  assembled matrix (cf.  grid.m assemblepoisson()), so perhaps this
  referee is just confused about this point.  Is the matrix assembled
  or is everything matrix free?  And how is the cost compared?}
\noindent
Since we employ hexahedral elements\footnote{For tetrahedral elements,
  this difference might be less pronounced.} in combination with a
tensorial finite element basis, the effect of matrix-free operations
for higher-order elements can be quite significant in terms of
floating point operations, memory requirements, and actual run time:

\begin{itemize}
\item \emph{Memory requirements for assembled matrices:} For an order
  $p$, assembled element matrices are dense and of the size
  $(p+1)^3\times(p+1)^3$. For $p=9$, for instance, $(p+1)^3=1000$ and
  thus each element contributes $10^6$ entries to the assembled
  stiffness matrix, and each row in the matrix contains, on average,
  several 1000 nonzero entries. Thus, for high orders, memory becomes
  a significant issue.
\item \emph{Floating point operations for matrix-free versus assembled
  MatVec:} For hexahedral elements, the operation count for a
  tensorized matrix-free matvec is $\mathcal{O}(p^4)$ as opposed to
  $\mathcal{O}(p^6)$ for a fully assembled matrix \cite{orszag80}.
   Please note that this is only considering the dense elemental
  matrix, so this is true for the best SpMV implementation possible
  % \note{I know what you mean, but the last sentence is still a bit confusing}. 
    This has been clarified and cited in the paper as
  well.
\end{itemize}
We have better incorporated these arguments into the paper, and refer
to \cite{orszag80,DevilleFischerMund02,ptatin3d} for floating point operation estimates
and memory access arguments, and to \cite{ptatin3d,BursteddeGhattasGurnisEtAl08} for
experiments and comparisons between matrix-free and assembled matvecs
on modern hardware. We acknowledge that results depend on $p$ (for
$p\le 2 or 3$, a MatVec based on an assembled matrix can be faster),
on the spatial dimension (2 vs.\ 3), on the equation (scalar
vs.\ vector system), the geometry deformation of the mesh (high-order
vs.\ low-order geometry transformation), and on the hardware (GPGPU,
CPU, etc.). However, to keep our comparison simple and avoid mixing
too many factors, we decided to use the fine-grid MatVecs as our
``performance measure''.\\[1ex]


\noindent
No, AMG was not tested. The text in the paper has been removed.\\[1ex]

\noindent
Yes, in the current MATLAB implementation, the matrix is
assembled. This is the reason we are comparing the iteration counts
(which is proportional to the number of matvecs).

\referee{ Figure 2 needs some work.  Column 1 displays the smoothing
  properties.  But why should we expect Jacobi, Chebyshev, and SSOR to
  display classic smoothing properties for high-order problems?  And
  why does weighted Jacobi annihilate all high-order modes (for
  linears) in the top corner plot?  In addition, the use of a single
  coarse grid is used here (so it should be labeled two-grid rather
  than V-cycle, but this is minor), but what is the coarse grid?  For
  presentation, a suggestion is to move the rather lengthy paragraph
  in the Figure 2 caption to the section where it is cited.
}
\noindent
We do \emph{not} expect these iterative methods to display classical
smoothing properties; their performance and robustness is actually
what we try to study numerically. 
% \todo{Check Ronquist/Patera 87 if  they can prove something in 1D.}

Weighted Jacobi does not annihilate the high-order error components
completely, but these error components are decreased so strongly
that they are out of the range of the $y$-axis (note that we are using
a logarithmic scale for the error).\\[1ex]
Thanks for the suggestion to replace ``v-cycle'' by ``two-grid
correction''.
%\todo{Need to actually do that in the paper}\\[1ex]
%\note{GS: I'm fine with moving some of the text in the caption to the section.}


\referee{
To use Cheychev, are the Arnoldi iterations considered in the cost?}
\noindent
No, they are not included in the cost. As argued earlier, they add 10 matvecs
to the cost. This has been clarified in the text.

\referee{ Algorithm 4.1 requires $M$ to be symmetric.  Are the
  smoothers applied symmetrically?  SSOR does not appear to do this in
  particular.}
\noindent
Yes, all the smoothers are symmetrically applied, including Symmetric-SOR.

\referee{ Section 4 gets very long without reporting anything on
  relative efficiency.  The iteration counts show that the methods are
  converging in a reasonable fashion.  But " As a consequence, the
  iteration numbers reported in the next section can be used to
  compare the efficiency of the different methods." on page 12 is as
  close as it gets to comparing the total work of the method(s).  The
  introduction reports this as a major contribution and so it should
  be central in section 4.
}
We have rewritten this section to make this clear, as well as the arguments
for the comparison based on the iteration counts. 

\referee{
A Few Minor Issues\\[.3ex]
page 3: discretization $\rightarrow$ discretizations}
\noindent
Thanks--has been corrected.
\referee{
page 3: on the performance $\rightarrow$ in the performance }
\noindent
Corrected.

\referee{
page 3: The self references to 15-17 are unnecesary as presented.  If
the intent is to motivate high-order, then other references should be
used.  This happens again on page 5 in section 2.3 where [15-16] are
mentioned with little connection to the current work.}
\noindent
We have removed these references.
\referee{
page 10: "number of multigrid v-cycles or of CG iterations" needs a
rewording
}
\noindent
This sentence has been rephrased.
\referee{ page 12: Table I should list these values as iterations
  counts somewhere (and there is a hanging sentence below the table
  from the itemize on the previous page) }
\noindent
The table captions have been modified to indicate the iteration counts.


\bibliographystyle{unsrt}
\bibliography{../ccgo,../mg}


\end{document}
